Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-07-03 paper2.qmd
review and editing

---
title: "Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation"
author:
  - name: 
      given: Jose Manuel
      family: Rivera Espejo
    orcid: 0000-0002-3088-2783
    url: https://www.uantwerpen.be/en/staff/jose-manuel-rivera-espejo_23166/
    email: JoseManuel.RiveraEspejo@uantwerpen.be
    corresponding: true
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Tine
      family: van Daal
      non-dropping-particle: van
    orcid: 0000-0001-9398-9775
    url: https://www.uantwerpen.be/en/staff/tine-vandaal/
    email: tine.vandaal@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Sven
      family: De Maeyer
      non-dropping-particle: De
    orcid: 0000-0003-2888-1631
    url: https://www.uantwerpen.be/en/staff/sven-demaeyer/
    email: sven.demaeyer@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Steven
      family: Gillis
    orcid: 
    url: https://www.uantwerpen.be/nl/personeel/steven-gillis/
    email: steven.gillis@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Linguistics
        group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
funding: 
  statement: "The project was founded through the Research Fund of the University of Antwerp (BOF)."
keywords:
  - causal inference
  - directed acyclic graphs
  - structural causal models
  - bayesian statistical methods
  - thurstonian model
  - comparative judgement
  - probability
  - statistical modeling
abstract: |
  This study revisits Thurstone's law of comparative judgment (CJ), focusing on two prominent issues of traditional approaches. First, it critiques the heavy reliance on Thurstone's Case V assumptions and, by extension, the Bradley-Terry-Luce (BTL) model when analyzing CJ data. Specifically, the study raises concerns about the assumptions of equal discriminal dispersions and zero correlation between the stimuli. While these assumptions simplify the trait measurement model, they may fail to capture the complexity of CJ data, potentially leading to unreliable and inaccurate trait estimates. Second, the study highlights the apparent disconnect between CJ's trait measurement and hypothesis testing processes. Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical results derived from these processes.

  To address these issues, the study extends Thurstone's general form using a systematic and integrated approach based on Causal and Bayesian inference methods. This extension integrates core theoretical principles alongside key CJ assessment design features, such as the selection of judges, stimuli, and comparisons. It then translates these elements into a probabilistic statistical model for analyzing dichotomous CJ data, overcoming the rigid assumptions of Case V and the BTL model.

@@

<!-- 1. Overview of CJ method in assessing traits -->
In *comparative judgment* (CJ) studies, judges assess a specific trait or attribute across different stimuli by performing pairwise comparisons [@Thurstone_1927a; @Thurstone_1927b]. Each comparison produces a dichotomous outcome, indicating which stimulus is perceived to have a higher trait level. For example, when assessing writing quality, judges compare pairs of written texts (the stimuli) to determine the relative writing quality each text exhibit (the trait) [@Pollitt_2012b; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

<!-- 2. What is the effectiveness of CJ -->
Numerous studies have documented the effectiveness of CJ in assessing traits and competencies over the past decade. These studies have highlighted three aspects of the method's effectiveness: its reliability, validity, and practical applicability. Research on reliability suggests that CJ requires a relatively modest number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to generate trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023]. In addition, the evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021]. Meanwhile, research on validity indicates the capacity of CJ scores to represent accurately the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023]. Lastly, research on its practical applicability highlights CJ's versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020b; @Boonen_et_al_2020].

@@

<!-- The A for ‘adaptivity’ in ACJ means that the choice of which objects are presented to the judges depends on the outcomes of judgments made so far – the idea being to use judge time efficiently by not presenting objects to compare where the result of the comparison is almost certain. [@Bramley_2015, pp. 4].  -->

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved. This study identifies and discusses two prominent issues of traditional approaches that can undermine the reliability and validity of CJ's trait estimates. First, it critiques the heavy reliance on Thurstone's Case V assumptions [@Thurstone_1927b] and, by extension, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] when analyzing CJ data. Specifically, the study raises concerns about the assumptions of equal discriminal dispersions and zero correlation between the stimuli. While these assumptions simplify the trait measurement model, they may fail to capture the complexity of some traits or account for heterogeneous stimuli, potentially leading to unreliable and inaccurate trait estimates. Second, the study highlights the disconnect between CJ's trait measurement and hypothesis testing processes. Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical inferences derived from these processes [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].

<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability. [@Perron_et_al_2015, pp. 2] -->

<!-- 5. So, what are we doing? -->
To address these issues, this study extends Thurstone's general form through a systematic and integrated approach that combines causal and Bayesian inference methods. In addition to potentially enhancing measurement reliability and validity, and improving statistical accuracy in hypothesis testing, this approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ assessments. Second, it shifts the current comparative data analysis paradigm from passively accepting Case V and the BTL model assumptions to actively testing whether those assumptions fit the data under analysis.

<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability. [@Perron_et_al_2015, pp. 2] -->

<!-- Many cross-cultural scientists already pay close attention to concerns of causal inference and comparison without use of a formal framework (e.g., Norenzayan & Heine, 2005; Pollet et al., 2014). For these researchers, a formalized framework can provide a vocabulary to articulate their concerns and work toward solutions in a more systematic manner. [@Deffner_et_al_2022, pp. 2] -->

<!-- Specifically, a framework is needed which lets us reason about the causes of non-invariance. [@Sterner_et_al_2024, pp. 751] -->

<!-- 4. what is the organization of the study? -->
The study divides its content into six main sections. @sec-thurstone_theory provides an overview of Thurstone's theory. @sec-theory-issues discusses the identified issues in detail. @sec-theoretical extends Thurstone's general form to address these challenges. The extension integrates core theoretical principles alongside key CJ assessment design features, such as the selection of judges, stimuli, and comparisons. @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model to analyze dichotomous pairwise comparison data. Finally, @sec-discussion discusses the findings, explores avenues for future research, and detail the challenges for future researchers.

@@

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone’s theory addresses pairwise comparisons wherein a single judge evaluates multiple stimuli [@Thurstone_1927b]. The theory posits that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* captures the psychological impact each stimulus exerts on the judge or, more simply, his perception of the stimulus trait. The theory assumes that the discriminal process for any given stimulus forms a Normal distribution along the trait continuum [@Thurstone_1927b]. The mode (mean) of this distribution, known as the *modal discriminal process*, indicates the stimulus position on this continuum, while its dispersion, referred to as the *discriminal dispersion*, reflects variability in the perceived trait of the stimulus.

<!-- The object of this paper is to describe a new psychophysical law which may be called the law of comparative judgment and to show some of its special applications in the measurement of psychological values. [@Thurstone_1927b, pp. 266] -->

<!-- It applies fundamentally to the judgments of a single observer who compares a series of stimuli by the method of paired comparison when no 'equal' judgments are allowed [@Thurstone_1927b, pp. 267] -->

<!-- The psychological continuum or scale is so constructed or defined that the frequencies of the respective discriminal processes for any given stimulus form a normal distribution on the psychological scale. This involves no assumption of a normal distribution or of anything else. The psychological scale is at best an artificial construct. If it has any physical reality, we certainly have not the remotest idea of what it may be like. ... We define the scale in terms of the frequencies of the discriminal processes for any stimulus. This artificial construct, the psychological scale, is so spaced off that the frequencies of the discriminal processes for any given stimulus form a normal distribution on the scale.[@Thurstone_1927b, pp. 266] -->

<!-- 2. Example of discriminal processes -->
@fig-discriminal_process illustrates the hypothetical discriminal processes along a quality trait continuum for two written texts. The figure indicates that the modal discriminal process for Text B is positioned further along the continuum than that of Text A $(T_{B} > T_{A})$, suggesting that Text B exhibits higher quality. Additionally, the figure highlights that Text B has a broader distribution compared to Text A, which arises from its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.

@@

![Discriminal processes](/images/png/discriminal_process.png){#fig-discriminal_process width=100%}

![Discriminal difference](/images/png/discriminal_difference.png){#fig-discriminal_difference width=100%}

@@

<!-- 3. Thurstone's Theory of CJ: discriminal difference -->
However, since the individual discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*. This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008]. This suggests that pairwise comparison outcomes depend on the relative distance between stimuli, not their absolute positions on the continuum. Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*, determines the observed dichotomous outcome. Furthermore, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978]. In this distribution, the mode (mean) represents the average relative separation between the stimuli, and its dispersion indicates the variability of that separation.

@@

<!-- The discriminal mode and dispersion corresponding to a single stimulus are inaccessible to observation. They can only be estimated when two objects are compared. Thurstone assumed that when two objects are compared with respect to a specified attribute, the object evoking the discriminal process further along the psychological continuum would be judged as possessing more of the attribute. ... The outcome of the paired comparison judgement is therefore related to the distribution of the difference between the two distributions of discriminal processes for essay A and essay B. [@Bramley_2008, pp. 251] -->

<!-- When subject v compares two statements, j and i, he/she reports that item j reflects more of the property of D_{vj}, - D_{vi}, > 0. Since D_{vj} and D_{vi}, are assumed normally distributed, the difference process ... is normally distributed with mean a_{j} - a_{i}, and variance s^{2}_{ji} [@Andrich_1978, pp. 452] -->

@@

<!-- 1. What is the problem with the general form -->
Thurstone noted from the outset that his general formulation, described in @sec-thurstone_theory, led to a *trait scaling problem*. Specifically, the model required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b]. For instance, in a CJ assessment with five texts, the general form would require estimating $20$ parameters: five modal discriminal processes, five discriminal dispersions, and $10$ correlations--one per comparison (see @tbl-thurstone_cases). However, a single judge could only provide ${5 \choose 2} = 10$ unique comparisons, an insufficient data set to estimate the required parameters.

<!-- According to this equation every pair of stimuli presents the possibility of a different correlation between the discriminal deviations. If this degree of freedom is allowed, the problem of psychological scaling would be insoluble because every observation equation would introduce a new unknown and the number of unknowns would then always be greater that the number of observation equations. In order to make the problem soluble it is necessary to make at least one assumption, namely that the correlation between discriminal deviations is practically constant throughout the stimulus series and for the single observer. [@Thurstone_1927b, pp. 267] -->

@@

<!-- 2. Thurstone's Cases and their simplifying assumptions -->
To address this issue and facilitate the practical implementation of the theory, Thurstone developed five cases derived from this general form, each progressively incorporating additional simplifying assumptions [@Thurstone_1927b]. In Case I, Thurstone postulated that pairs of stimuli would maintain a constant correlation across all comparisons. In Case II, he allowed multiple judges to undertake comparisons instead of confining evaluations to a single judge. In Case III, he posited that there was no correlation between stimuli. In Case IV, he assumed that the stimuli exhibited similar dispersions. Finally, in Case V, he replaced this assumption with the condition that stimuli had equal discriminal dispersions. @tbl-thurstone_cases summarizes the assumptions of the general form and the five cases. For a detailed discussion of these cases and their progression, refer to @Thurstone_1927b and @Bramley_2008.

<!-- For the practical application of the law of comparative judgment we shall consider five cases which differ in assumptions, approximations, and degree of simplification. The more assumptions we care to make, the simpler will be the observation equations. [@Thurstone_1927b, pp. 267] -->

<!-- In Case II, Thurstone assumes that "each judge compares each stimulus with every other stimulus only once" [@Thurstone_1927b, pp. 268]. -->

<!-- 3. But are our specific problems? -->
However, Thurstone developed Case V prioritizing statistical simplicity over precise trait measurement and offering no guidance on how to use its trait estimates for statistical inference or hypothesis testing. Specifically, Thurstone cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], as it imposes the most extensive set of simplifying assumptions [@Bramley_2008; @Kelly_et_al_2022] (see @tbl-thurstone_cases). Moreover, because Thurstone's primary goal was to produce a "rather coarse scaling" of traits and "allocate the compared stimuli on this continuum" [@Thurstone_1927b, pp. 269], his theory did not support formal statistical inference. Despite these limitations, CJ research has predominantly relied on Case V to measure different traits, which raises significant concerns about the reliability and validity of such measurements in contexts where the case's assumptions may not hold [@Kelly_et_al_2022; @Andrich_1978]. Furthermore, although the CJ tradition has attempted to address the gap in hypothesis testing by relying on the point estimates of traits--or their transformations--the statistical literature cautions against using these estimates as the sole basis for statistical inference, as such practices introduce bias in the analysis and reduce the precision of hypothesis tests [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023]. Next, both issues are discussed in more depth.

<!-- Case V - This is the simplest formulation of the law and it involves, in addition to the previous assumptions, the assumption that all the discriminal dispersions are equal. This assumption should not be made without experimental test. [@Thurstone_1927b, pp. 270] -->

<!-- Case 5 makes the greatest simplification by assuming that the discriminal dispersions are all equal [@Bramley_2008, pp. 253] -->

<!-- Adding assumptions simplifies the formula and improves its tractability, but one cannot place as much confidence in the outputs of the formula if those assumptions are not met [@Kelly_et_al_2022, pp. 677]. -->

<!-- This is a simple observation equation which may be used for rather coarse scaling . .... This law defines the psychological scale or continuum. It allocates the compared stimuli on the continuum [@Thurstone_1927b, pp. 269] -->

<!-- Consequently, different samples, whether drawn randomly or otherwise from the same empirical population, should give the same relative scale value estimates. In this sense the models are also "sample-free." 
However, this property of the model is true in empirical data only if the model holds in the data. Therefore, "sample-freeness" is always only a hypothesis until checked, as is the assumption that the model holds. [@Andrich_1978, pp.459] -->

@@

<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
As previously discussed, Case V remains the most widely used model in CJ literature. This preference largely stems from the widespread adoption of the BTL model, which provides a simplified statistical representation of the case. The BTL model mirrors most of Case V's assumptions, with one notable distinction. While Case V assumes a Normal distribution for the stimuli' discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008] (see @tbl-thurstone_cases). However, this substitution has minimal impact on trait estimation or model interpretation because the scale of the discriminal process (i.e., the latent trait) is arbitrary up to a non-monotonic transformation [@vanderLinden_et_al_2017_I; @McElreath_2021]. That is, as long as the substitution (transformation) preserves the data rank order, the choice of distribution for the discriminal processes is inconsequential. This condition is satisfied in this case, as the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I].

@@

<!-- An alternative to the normal density function is the logistic. The substitution of the logistic for the normal is commonly made in circumstances such as the above (cf. Birnbaum, 1968; Bock, 1975) because it is more tractable. [@Andrich_1978, pp. 455] -->

>>>>

---
title: "Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation"
author:
  - name: 
      given: Jose Manuel
      family: Rivera Espejo
    orcid: 0000-0002-3088-2783
    url: https://www.uantwerpen.be/en/staff/jose-manuel-rivera-espejo_23166/
    email: JoseManuel.RiveraEspejo@uantwerpen.be
    corresponding: true
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Tine
      family: van Daal
      non-dropping-particle: van
    orcid: 0000-0001-9398-9775
    url: https://www.uantwerpen.be/en/staff/tine-vandaal/
    email: tine.vandaal@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Sven
      family: De Maeyer
      non-dropping-particle: De
    orcid: 0000-0003-2888-1631
    url: https://www.uantwerpen.be/en/staff/sven-demaeyer/
    email: sven.demaeyer@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Steven
      family: Gillis
    orcid: 
    url: https://www.uantwerpen.be/nl/personeel/steven-gillis/
    email: steven.gillis@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Linguistics
        group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
funding: 
  statement: "The project was founded through the Research Fund of the University of Antwerp (BOF)."
keywords:
  - causal inference
  - directed acyclic graphs
  - structural causal models
  - bayesian statistical methods
  - thurstonian model
  - comparative judgement
  - probability
  - statistical modeling
abstract: |
  This study revisits Thurstone's law of comparative judgment (CJ) by addressing two central issues in the CJ literature. First, it critiques the widespread reliance on Thurstone's Case V assumptions and, by extension, the Bradley-Terry-Luce (BTL) model in the analysis of CJ data. Specifically, this study argues that while the assumptions of equal discriminal dispersions and zero correlation between stimuli simplify the trait measurement model, they may fail to capture the complexity of some traits or account for heterogeneous stimuli, potentially leading to unreliable and inaccurate trait estimates. Second, the study highlights the apparent disconnect between trait measurement and hypothesis testing procedures in CJ applications. It contends that while separating these procedures simplifies the analysis of CJ data, this practice can also undermine the resulting statistical inferences.

  To address these issues, this study aims to extend Thurstone's general form using a systematic and integrated approach based on causal and Bayesian inference methods. This extension integrates Thurstone's core theoretical principles alongside key CJ assessment design features. It then translates these elements into a probabilistic statistical model for analyzing dichotomous CJ data, overcoming the rigid assumptions of Case V and the BTL model.

@@

<!-- 1. Overview of CJ method in assessing traits -->
In *comparative judgment* (CJ) studies, judges assess a specific trait or attribute across different stimuli by performing pairwise comparisons [@Thurstone_1927a; @Thurstone_1927b]. Each comparison produces a dichotomous outcome, indicating which stimulus is perceived to have a higher trait level. For example, judges can compare pairs of written texts (the stimuli) to determine the relative writing quality each text exhibit (the trait) [@Pollitt_2012b; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

<!-- 2. What is the effectiveness of CJ -->
Numerous studies have documented the effectiveness of CJ in assessing traits and competencies over the past decade. These studies have highlighted three key strengths of the method: its reliability, its validity, and practical applicability. Research on reliability suggests that CJ requires a relatively modest number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to generate trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023]. In addition, the evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021]. Meanwhile, research on validity indicates the capacity of CJ scores to represent accurately the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023]. Lastly, research on its practical applicability highlights CJ's versatility across educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020b; @Boonen_et_al_2020].

@@

<!-- The A for ‘adaptivity’ in ACJ means that the choice of which objects are presented to the judges depends on the outcomes of judgments made so far – the idea being to use judge time efficiently by not presenting objects to compare where the result of the comparison is almost certain. [@Bramley_2015, p. 4].  -->

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved. This study identifies and discusses two prominent issues in the CJ literature that can undermine the reliability and validity of CJ's trait estimates. First, it critiques the widespread reliance on Thurstone's Case V assumptions [@Thurstone_1927b] and, by extension, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] in the analysis of CJ data. Specifically, this study argues that while the assumptions of equal discriminal dispersions and zero correlation between stimuli simplify the trait measurement model, they may fail to capture the complexity of some traits or account for heterogeneous stimuli, potentially leading to unreliable and inaccurate trait estimates. Second, the study highlights the disconnect between trait measurement and hypothesis testing procedures in CJ applications. In particular, it contends that while separating these procedures simplifies the analysis of CJ data, this practice can also undermine the resulting statistical inferences [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].

<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability. [@Perron_et_al_2015, p. 2] -->

<!-- 5. So, what are we doing? -->
To address these issues, this study aims to extend Thurstone's general form through a systematic and integrated approach that combines causal and Bayesian inference methods. This extension integrates Thurstone's core theoretical principles alongside key CJ assessment design features, such as the selection of judges, stimuli, and comparisons. In addition to potentially enhancing measurement reliability and validity, and improving statistical accuracy in hypothesis testing, this approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ assessments. Second, it shifts the current comparative data analysis paradigm from passively accepting Case V and the BTL model assumptions to actively testing whether those assumptions fit the data under analysis.

<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability. [@Perron_et_al_2015, p. 2] -->

<!-- Many cross-cultural scientists already pay close attention to concerns of causal inference and comparison without use of a formal framework (e.g., Norenzayan & Heine, 2005; Pollet et al., 2014). For these researchers, a formalized framework can provide a vocabulary to articulate their concerns and work toward solutions in a more systematic manner. [@Deffner_et_al_2022, p. 2] -->

<!-- Specifically, a framework is needed which lets us reason about the causes of non-invariance. [@Sterner_et_al_2024, p. 751] -->

<!-- 4. what is the organization of the study? -->
The remainder of this study is organized into seven sections. @sec-thurstone_theory provides an overview of Thurstone's theory. @sec-theory-issues examines the central issues identified in the CJ literature. @sec-theoretical extends Thurstone's general form to address these issues. The extension integrates core theoretical principles alongside key CJ assessment design features, such as the selection of judges, stimuli, and comparisons. @sec-statistical translates these theoretical and design elements into a probabilistic statistical model to analyze dichotomous pairwise comparison data. @sec-discussion reviews the findings, outline directions for future research, discusses the study's limitations and details the challenges that applied researchers may encounter. Finally, @sec-conclusion provides the concluding remarks.

@@

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its *general form*, Thurstone’s theory addresses pairwise comparisons of a single judge who evaluates multiple stimuli [@Thurstone_1927a; @Thurstone_1927b]. The theory posits that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* captures the psychological impact each stimulus exerts on the judge or, more simply, his perception of the stimulus trait. The theory assumes that the discriminal process for any given stimulus forms a Normal distribution along the trait continuum [@Thurstone_1927b]. The mode (mean) of this distribution, known as the *modal discriminal process*, indicates the stimulus position on this continuum, while its dispersion, referred to as the *discriminal dispersion*, reflects variability in the perceived trait of the stimulus.

<!-- The object of this paper is to describe a new psychophysical law which may be called the law of comparative judgment and to show some of its special applications in the measurement of psychological values. [@Thurstone_1927b, p. 266] -->

<!-- It applies fundamentally to the judgments of a single observer who compares a series of stimuli by the method of paired comparison when no 'equal' judgments are allowed [@Thurstone_1927b, p. 267] -->

<!-- The psychological continuum or scale is so constructed or defined that the frequencies of the respective discriminal processes for any given stimulus form a normal distribution on the psychological scale. This involves no assumption of a normal distribution or of anything else. The psychological scale is at best an artificial construct. If it has any physical reality, we certainly have not the remotest idea of what it may be like. ... We define the scale in terms of the frequencies of the discriminal processes for any stimulus. This artificial construct, the psychological scale, is so spaced off that the frequencies of the discriminal processes for any given stimulus form a normal distribution on the scale.[@Thurstone_1927b, p. 266] -->

<!-- 2. Example of discriminal processes -->
These ideas become clearer with an example. @fig-discriminal_process illustrates the hypothetical discriminal processes for two written texts along a *quality* trait continuum. The figure shows that the modal discriminal process for Text B lies further along the continuum than that of Text A $(T_{B} > T_{A})$, suggesting that Text B exhibits higher quality. The figure also shows that Text B has a broader distribution than Text A, indicating a greater variability in perception due to its larger discriminal dispersion $(\sigma_{B} > \sigma_{A})$.

@@

![Hypothetical discriminal processes for two written texts](/images/png/discriminal_process.png){#fig-discriminal_process width=100%}

![Hypothetical discriminal difference of two written texts](/images/png/discriminal_difference.png){#fig-discriminal_difference width=100%}

@@

<!-- 3. Thurstone's Theory of CJ: discriminal difference -->
However, since the individual discriminal processes of the stimuli are not directly observable [@Thurstone_1927a], the theory introduces the *law of comparative judgment*. This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008]. This suggests that pairwise comparison outcomes depend on the relative distance between stimuli, not their absolute positions on the continuum. Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*, determines the observed dichotomous outcome. Furthermore, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978]. In this distribution, the mode (mean) represents the average relative separation between the stimuli $(T_{B} - T_{A})$, and its dispersion $\sigma_{BA}$ indicates the variability of that separation.

@@

<!-- The discriminal mode and dispersion corresponding to a single stimulus are inaccessible to observation. They can only be estimated when two objects are compared. Thurstone assumed that when two objects are compared with respect to a specified attribute, the object evoking the discriminal process further along the psychological continuum would be judged as possessing more of the attribute. ... The outcome of the paired comparison judgement is therefore related to the distribution of the difference between the two distributions of discriminal processes for essay A and essay B. [@Bramley_2008, p. 251] -->

<!-- When subject v compares two statements, j and i, he/she reports that item j reflects more of the property of D_{vj}, - D_{vi}, > 0. Since D_{vj} and D_{vi}, are assumed normally distributed, the difference process ... is normally distributed with mean a_{j} - a_{i}, and variance s^{2}_{ji} [@Andrich_1978, p. 452] -->

@@

<!-- 1. What is the problem with the general form -->
Thurstone noted from the outset that his general formulation, described in @sec-thurstone_theory, led to a *trait scaling problem*. Specifically, the model required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b]. For instance, in a CJ assessment with five texts, the general form would require estimating $20$ parameters: five modal discriminal processes, five discriminal dispersions, and $10$ correlations--one per comparison. However, a single judge could only provide ${5 \choose 2} = 10$ unique comparisons, an "insufficient" data set to estimate the required parameters.

<!-- According to this equation every pair of stimuli presents the possibility of a different correlation between the discriminal deviations. If this degree of freedom is allowed, the problem of psychological scaling would be insoluble because every observation equation would introduce a new unknown and the number of unknowns would then always be greater that the number of observation equations. In order to make the problem soluble it is necessary to make at least one assumption, namely that the correlation between discriminal deviations is practically constant throughout the stimulus series and for the single observer. [@Thurstone_1927b, p. 267] -->

<!-- 2. Thurstone's Cases and their simplifying assumptions -->
To address this issue and facilitate the practical implementation of the theory, Thurstone developed five cases derived from this general form, each progressively incorporating additional simplifying assumptions [@Thurstone_1927b]. In Case I, Thurstone postulated that pairs of stimuli would maintain a constant correlation across all comparisons. In Case II, he allowed multiple judges to undertake comparisons instead of confining evaluations to a single judge. In Case III, he posited that there was no correlation between stimuli. In Case IV, he assumed that the stimuli exhibited similar dispersions. Finally, in Case V, he replaced this assumption with the condition that stimuli had equal discriminal dispersions. @tbl-thurstone_cases summarizes the assumptions of the general form and the five cases. For a detailed discussion of these cases and their growing simplification, refer to @Thurstone_1927b and @Bramley_2008.

@@

<!-- For the practical application of the law of comparative judgment we shall consider five cases which differ in assumptions, approximations, and degree of simplification. The more assumptions we care to make, the simpler will be the observation equations. [@Thurstone_1927b, p. 267] -->

<!-- In Case II, Thurstone assumes that "each judge compares each stimulus with every other stimulus only once" [@Thurstone_1927b, p. 268]. -->

<!-- 3. But are our specific problems? -->
As the table suggest, Thurstone developed Case V with an emphasis on statistical simplicity, but this simplicity comes at the expense of accurate and precise trait measurement and practical guidance for inference. Specifically, Thurstone cautioned that Case V use "should not be made without (an) experimental test" [@Thurstone_1927b, p. 270], as it imposes the most extensive set of simplifying assumptions [@Bramley_2008; @Kelly_et_al_2022] (see @tbl-thurstone_cases). Moreover, because Thurstone's primary goal was to produce a "rather coarse scaling" of traits and "allocate the compared stimuli on this continuum" [@Thurstone_1927b, p. 269], his theory did not support formal statistical inference. Nevertheless, despite Case V's limitations, CJ research has predominantly relied on it to measure various traits, raising significant concerns about the reliability and validity of such measurements in contexts where its assumptions may not hold [@Kelly_et_al_2022; @Andrich_1978]. Furthermore, although the CJ tradition has attempted to address the gap in statistical inference by relying on the point estimates of traits--or their transformations--the statistical literature cautions against using these estimates as the sole basis for inference, as such practices introduce bias and reduce the precision of hypothesis tests [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023]. The next sections discuss both issues in greater depth.

<!-- Case V - This is the simplest formulation of the law and it involves, in addition to the previous assumptions, the assumption that all the discriminal dispersions are equal. This assumption should not be made without experimental test. [@Thurstone_1927b, p. 270] -->

<!-- Case 5 makes the greatest simplification by assuming that the discriminal dispersions are all equal [@Bramley_2008, p. 253] -->

<!-- Adding assumptions simplifies the formula and improves its tractability, but one cannot place as much confidence in the outputs of the formula if those assumptions are not met [@Kelly_et_al_2022, p. 677]. -->

<!-- This is a simple observation equation which may be used for rather coarse scaling . .... This law defines the psychological scale or continuum. It allocates the compared stimuli on the continuum [@Thurstone_1927b, p. 269] -->

<!-- Consequently, different samples, whether drawn randomly or otherwise from the same empirical population, should give the same relative scale value estimates. In this sense the models are also "sample-free." 
However, this property of the model is true in empirical data only if the model holds in the data. Therefore, "sample-freeness" is always only a hypothesis until checked, as is the assumption that the model holds. [@Andrich_1978, p.459] -->

@@

<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
Case V is the most widely used model in the CJ literature. This preference largely stems from the widespread adoption of the BTL model, which provides a simplified statistical representation of the case. The BTL model incorporates most of Case V's assumptions, with one notable exception. While Case V assumes a Normal distribution for the stimuli' discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008] (see @tbl-thurstone_cases). However, this substitution has minimal impact on trait estimation or model interpretation because the scale of the discriminal process (i.e., the latent trait) is arbitrary up to a non-monotonic transformation [@vanderLinden_et_al_2017_I; @McElreath_2021]. In other words, as long as the substitution (transformation) preserves the rank order of the data, the choice of distribution for the discriminal processes is inconsequential. The BTL model satisfies this condition because the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I].

@@

<!-- An alternative to the normal density function is the logistic. The substitution of the logistic for the normal is commonly made in circumstances such as the above (cf. Birnbaum, 1968; Bock, 1975) because it is more tractable. [@Andrich_1978, p. 455] -->
